---
title: "Gapminder_exercise"
author: "Rikke Uldbæk"
date: "10/31/2022"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

## set working directory
setwd("~/Desktop/OneDrive - Aarhus Universitet/Cognitive Science/5th semester/Cultural Datascience/AU674347_uldbaek_rikke/6.W43.Practicing functions with Gapminder")
```

#### Loading the data and packages
```{r}
#packages
pacman::p_load(tidyverse, ggplot2, gapminder, stringr)
library(gapminder)

#loading data from gapminder package
df <- gapminder
df <- as.data.frame(df)
```


## 1) Define a Function
Define a defensive function that calculates the Gross Domestic Product of a nation from the data available in the gapminder dataset. You can use the population and GDPpercapita columns for it. Using that function, calculate the GDP of Denmark in the following years: 1967, 1977, 1987, 1997, 2007, and 2017.


```{r}
#Define a function to calculate GDP of a nation for a given year, using the columns "population" and "GDPpercapita"
GDP_func <- function(data, chosen_nation, chosen_year) { #function arguments
  
  #isolate specified year and country
  data <- data[data$country %in% chosen_nation, ]
  data <- data[data$year %in% chosen_year, ]

  #calculate GDP
  GDP <- data$pop * data$gdpPercap
  data$GDP <- GDP

  return(data)
}

# function works fine
GDP_func(df, "Bulgaria", 1967)
```

Applying the function on Denmark 

```{r}
#Using the function on Denmark
dk_years <- c(1967, 1977, 1987, 1997, 2007, 2017)

for (i in dk_years){
  if(i %in% df$year){
    dk_GDP <- GDP_func(df, "Denmark", i)
    print(dk_GDP)
  }else{
    print(paste0("data was not found for year: ",i ))
  }
}
```



## 2) Make a Loop
Write a script that loops over each country in the gapminder dataset, tests whether the country starts with a ‘B’ , and prints out whether the life expectancy is smaller than 50, between 50 and 70, or greater than 70. (Hint: remember the grepl function, and review the Control Flow tutorial)

```{r}
# creating an initial dataframe
b_df <- data.frame()

# loop 1
for(i in unique(df$country)) {
  if(grepl("^B", i)){
    b_df <- rbind(b_df, df[which(df$country %in% i),]) }
  
}

# loop 2
for(j in unique(b_df$country)){
  
  new_df <- b_df %>% 
    filter(country == j)
  
  # Calculation of mean life expectancy
  mean <- mean(new_df$lifeExp)
    
  # Output of each B-country and their mean life expectancy and
  # whether they have a long-, short-, and "medium" life expectancy 
   if (mean > 70) { 
     print(paste0(j, " has a long life expectancy above 70 years")) 
    } else {
      if (mean < 50) { 
       print(paste0(j, " has a short life expectancy below 50 years")) 
     } else {
      print(paste0(j, " has a life expectancy between 50 and 70 years of age")) 
    }
  }
}
```
*For some reason, probably my poor coding skills, I could not merge loop 1 and loop 2. However, I solved the exercise anyway, and allowed my self to calculate the mean of life expectancy for each county (starting with B) over the years, to minimize the output and make it more manageable.*




## 3) Optional
*I choose not to make this exercise.*

